---
title: "Introduction to statistics in R: Day 1"
author: "Marilin Moor"
date: "2026-06-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to statistics in R: Day 1

Day 1 focus: exploratory data analysis (EDA) and two-group comparisons.

```{r}
# imports
library(tidyverse)
library(patchwork)
library(dplyr)
library(ISwR)    
library(ggplot2)
library(ggpubr)
library(rstatix)
```

### 1. Exploratory data analysis (EDA)

For revising EDA, we will use the `hellung` dataset (from `ISwR`):
describing growth conditions of the *Tetrahymena* unicellular organism:
cell `conc` (cells/mL), `diameter` (µm), and `glucose` (1 = with
glucose, 2 = without).

```{r}
?hellung
data(hellung)
head(hellung)
```

#### Descriptive statistics

```{r}
summary(hellung)
mean(hellung$conc)
```

**Task 1**: Provide at least two ways to compute the mean of cell
concentration.

```{r}
mean(hellung$conc)

summary(hellung$conc)["Mean"]

hellung %>% summarise(mean_conc = mean(conc))
```

**Task 2**: Using `dplyr`, find the average diameter of cells grown
*with* glucose.

```{r}
# hint: filter() then summarise()
hellung %>% 
  filter(glucose == 1) %>%
  summarise(mean_dm = mean(diameter))
```

**Task 3**: Show the distributions of all numeric features in separate
plots and combine them in one frame using `patchwork`. See
<https://www.data-to-viz.com/> for plotting inspiration.

```{r}
p_conc <- ggplot(hellung, aes(conc)) +
  geom_histogram(fill = "blue")

p_diam <- ggplot(hellung, aes(diameter)) +
  geom_density(fill = "orange")

p_conc / p_diam
```

**Task 4**: Visualise whether glucose addition affects cell diameter,
and whether cell concentration is associated with diameter.

```{r}
# Does glucose affect diameter? 
ggplot(hellung, aes(x = factor(glucose), y= diameter)) +
  geom_boxplot()

# Does concentration relate to diameter?
ggplot(hellung, aes(x=conc, y=diameter)) + 
  geom_point() 
```

------------------------------------------------------------------------

### 2. Hypothesis testing - two-group comparisons

Theory slides: "slides/Introduction_to_hypothesis_testing_26.pdf"

We will focus on the following tests: the paired and two-sample t-tests
and their non-parametric alternatives. The one-sample t-test is shown
briefly at the end.

#### 2.1 Paired t-test - qPCR expression

*Use case*: "I measured the same samples under two conditions (control
vs. treated). Did the treatment change the response?"

Dataset: `data/qpcr_expression.csv` — ΔCt values for a target gene
normalised to a housekeeping gene, measured on the same biological
samples in a control and a treated condition.

```{r}
qpcr <- read.csv("data/qpcr_expression.csv")
head(qpcr)
```

**Task 1**: Visualise the paired data with a scatter of treated vs
control: each point is one sample. Add a linear line dashed red line
where `y = x`. (Points below the line mean the treatment lowered ΔCt (=
raised expression))

```{r}
ggplot(qpcr, aes(x=control_dCt, y=treated_dCt)) + 
  geom_abline(slope = 1, intercept =0, color = "red", linetype="dashed") +
  geom_point() 
```

**Assumptions of the paired t-test**

-   each pair of observations comes from the same subject/unit (study
    design);
-   the per-sample differences are approximately normally distributed.

**Task 2**: Compute the differences and check normality with a Q-Q plot
and Shapiro-Wilk test (`shapiro.test()`).

```{r}
diff <- qpcr$treated_dCt - qpcr$control_dCt
shapiro.test(diff)

ggplot(qpcr, aes(diff)) +
  geom_histogram(fill = "blue")
```

**Task 3**: Run the paired t-test and interpret the result.

```{r}
# t.test takes numeric vectors, rstatix: t_test takes a dataframe/tibble
t.test(x = qpcr$treated_dCt, y = qpcr$control_dCt, paired = TRUE)
```

**Task 4:** Make sense of the output by answering the following
questions:

1)  What sided test is this?

Answer: 2-sided

2)  What values does the 95 percent confidence interval refer to?

Answer:

3)  Describe how the t-statistic is created.

Answer:

4)  Describe how the p-value is derived. Reach the same value using R
    with using just t-statistic.

```{r}
# write code here 

2 * pt(-abs(-8.0653), df=17)
```

------------------------------------------------------------------------

#### 2.2 Two-sample t-test - catalyst yield

*Use case*: "I have two independent groups (A and B). Do they differ on
average?"

Dataset: `data/catalyst_yield.csv` shows reaction yield (%) measured for
four catalysts (A, B, C, D), 10 replicates each. In this task we will
use 2 catalysts, A and B (`ab` dataframe). We want to find out if the
catalysts' (A, B) effect on the reaction yield differs between each
other.

```{r}
yield_data <- read.csv("data/catalyst_yield.csv")
head(yield_data)

ab <- yield_data %>% filter(catalyst %in% c("A", "B"))

ggplot(ab, aes(x = catalyst, y = yield_pct, fill = catalyst)) +
  geom_boxplot() +
  geom_jitter() +
  labs(y = "Yield (%)", title = "Reaction yield: catalyst A vs B") +
  theme_minimal() + theme(legend.position = "none")
```

**Assumptions**

-   the yield in each group is approximately normally distributed;
-   the variances of the two groups are equal (homoscedasticity); use
    **Levene's test** (`levene_test()`).

**Task 1**: Find out if the assumptions above are met with the current
data set.

```{r}
# in each group A, B 
a_df <- ab %>%
  filter(catalyst == "A")

b_df <- ab %>%
  filter(catalyst == "B")

shapiro.test(a_df$yield_pct)
shapiro.test(b_df$yield_pct)

levene_test(ab, yield_pct ~ catalyst)
```

**Task 2**: Run the two-sample t-test and interpret it. If variances are
unequal, use Welch's t-test by setting `var.equal = FALSE` (the default
in R).

```{r}
t.test(yield_pct ~ catalyst, ab, var.equal = TRUE)
```

#### 2.3 One-sample t-test

*Use case*: "Does my sample mean differ from a known reference value?"
Uncommon as true population mean is rarely available.

Dataset: `intake` dataset (from `ISwR`) describing energy intake for 11
women.

```{r}
data(intake)
?intake
intake
```

**Task**: Test whether women's pre-menstrual energy intake differs from
a recommended value of 7725 kJ.

```{r}
t.test(intake$pre, mu = 7725)
```

Let's see what we have learned so far:
<https://www.menti.com/al5iszppvero>

------------------------------------------------------------------------

#### 2.4 Non-parametric alternatives

Theory slides: "slides/Parametric_and_non_parametric_tests_26.pdf"

When the normality assumption fails (small samples, skewed data), use:

-   **Mann-Whitney U test** (`wilcox.test() /``wilcox.test()`) —
    independent groups, analogous to the two-sample t-test;
-   **Wilcoxon signed-rank test**
    (`wilcox.test(..., paired = TRUE) /``wilcox.test(.., paired = TRUE)`)
    — paired data, analogous to the paired t-test;
-   **Data transformation** to fix skewed data so a t-test becomes
    appropriate again (but beware of the transformed values)

##### 2.4.1 Mann-Whitney U test

Dataset: We will be using the air quality data set for this task. We
want to determine whether ozone levels differ significantly between May
and July.

```{r}
data(airquality)
head(airquality)
```

```{r}
may_july <- airquality %>%
  filter(Month %in% c(5, 7)) %>%
  mutate(Month = factor(Month, labels = c("May", "July")))
```

**Task 1:** Plot a combined histogram for ozone levels in May and July.

```{r}
ggplot(may_july, aes(x = Ozone, fill = Month)) +
  geom_histogram()
```

**Task 2:** Remove NAs from the Ozone column. Check normality, perform
log transformation if needed.

```{r}
!is.na(may_july$Ozone)
```

```{r}
may_july_no_na <- may_july  %>%
  filter(!is.na(Ozone))

may_df <- may_july_no_na %>%
  filter(Month == "May")

july_df <- may_july_no_na %>%
  filter(Month == "July")

shapiro.test(may_df$Ozone)
shapiro.test(july_df$Ozone)

may_df$log_Ozone <- log(may_df$Ozone)
july_df$log_Ozone <- log(july_df$Ozone)

shapiro.test(may_df$log_Ozone)
shapiro.test(july_df$log_Ozone)
```

**Task 3:** Test the difference between the ozone levels in May and
July.

```{r}
wilcox.test(may_df$Ozone, july_df$Ozone)
```

##### 2.4.2 Wilcoxon signed-rank example

**Task**: A basketball coach tests a new training plan, measuring the
performance of 15 athletes before (`before`) and after (`after`) 4
weeks. Determine whether the plan has an increased effect on the
performance. Check the normality of the differences first, then decide
which test to use.

```{r}
before <- c(80,82,85,87,90,91,92,93,94,95,96,97,98,99,100)
after  <- c(79,81,84,86,89,90,91,92,93,94,95,96,97,98,85)

diffs <- after - before

shapiro.test(diffs)

#wilcox.test(before, after, paired = TRUE, alternative = "greater")
wilcox.test(after, before, paired = TRUE, alternative = "greater")
```

------------------------------------------------------------------------

### 3. Wrap-up of Day 1

-   EDA with `dplyr`, `ggplot2`, and `patchwork`.
-   Picking the right two-group test: paired vs. independent; parametric
    (t-test) vs. non-parametric (Wilcoxon / Mann-Whitney); with
    assumption checks (normality, equal variance).

**This marks the end of Day 1. Congrats!**

------------------------------------------------------------------------

## Appendix: Power analysis (self-study)

How many samples do you need? Use `pwr.t.test()` for a precise
calculation.

```{r}
if (!require(pwr)) 
  { install.packages("pwr"); library(pwr) }

cohens_d <- 0.5            # standardised effect size

pwr.t.test(d = cohens_d, power = 0.8, sig.level = 0.05, type = "two.sample")
pwr.t.test(d = cohens_d, power = 0.8, sig.level = 0.05,
           type = "two.sample", alternative = "greater")
```
