---
title: "Introduction to statistics in R — Day 2"
author: "Marilin Moor"
date: "2026-06-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to statistics in R: Day 2

```{r}
# imports
library(tidyverse)
library(patchwork)
library(dplyr)
library(ISwR)
library(ggplot2)
library(ggpubr)
library(rstatix)
```

### Recap

Yesterday we covered EDA and two-group comparisons (paired & two-sample
t-tests, their non-parametric alternatives, and one-sample t-test).
Let's open `stat_test_tree.html` to look at the options for tests.

Today we extend to:

1.  **Multiple testing correction**: needed whenever we run many tests.
2.  **Multiple-group comparisons**: ANOVA (mention multi-factorial),
    post-hoc Tukey, Kruskal-Wallis, Dunn.
3.  **Correlation & regression**: Pearson, Spearman, Kendall, linear
    regression.
4.  **Categorical data**: Chi-square, Fisher's exact, odds ratio.

------------------------------------------------------------------------

### 1. Multiple testing correction

When you run many tests against the same hypothesis, the chance of
getting at least one false positive grows.

$$P(\text{at least 1 FP}) = 1 - (1 - \alpha)^n$$

where n is the number of tests, and $$\alpha$$ is the significance
level.

What does this false positive mean? The false positive is a test result
that appears significant even though the null hypothesis is true.

What does multiple testing mean?

For 1 test you are willing to accept a 5% chance of incorrectly
declaring significance of a test.

20 tests would look something like having 20 two-sample t-tests for
comparing expression between control and treatment groups of 20
different genes. For 20 tests, this chance of at least one FP is risen
to 64%. Multiple testing can refer to having multiple different null
hypotheses.

2 main modes to control false positives:

FWER: probability of at least one false positive (stricter approach) -
"how often do I get any false positive?"

FDR: proportion of false positives among rejected H0-s (more lenient) -
"of all results I call significant, what fraction are expected to be
false positives?"

Different correction methods control the amount of FP-s you may get:

-   **Bonferroni**: significance threshold becomes: $\alpha$ / m, adjust
    p-value would be: p\*m (p=initial p-value, m = number of tests).
    Strict, controls FWER.
-   **Holm**: uses sorting and p-value adjusting to find the suitable
    threshold, less conservative than Bonferroni.
-   **Benjamini-Hochberg** uses sorting and slightly different p-value
    adjusting, controls FDR. Most permissive.

Below we simulate 8 tests, each with 40 values from normal distribution
(μ, 1) for different μ. A one-sample t-test (H0: μ = 0, H1:μ ≠ 0) is run
with each group.

```{r}
set.seed(123)

n <- 40

groups <- list(
  A = rnorm(n, mean = 0.8),   
  B = rnorm(n, mean = 0.4),   
  C = rnorm(n, mean = 0.1),   
  D = rnorm(n, mean = 0.0),   
  E = rnorm(n, mean = -0.1),  
  F = rnorm(n, mean = -0.6),  
  G = rnorm(n, mean = -1.0),  
  H = rnorm(n, mean = 0.3)    
)

p_values <- sapply(
  groups,
  function(group) {           # for each group
    t.test(group, mu = 0)$p.value  # run t-test and return p-value
  }
)

# this sapply() is equivalent to: 
#p_values <- c()
#for (group_name in names(groups)) {
#  group_data <- groups[[group_name]]
#  test_result <- t.test(group_data, mu = 0)
#  p_values[group_name] <- test_result$p.value
#}

data.frame(p_values)
```

Here we can see the effect of adjusting p-values with different methods.

```{r}
p_bonf  <- p.adjust(p_values, method = "bonferroni")
p_holm  <- p.adjust(p_values, method = "holm")
p_bh    <- p.adjust(p_values, method = "BH")  # Benjamini–Hochberg

results <- data.frame(
  raw_p = p_values,

  bonferroni_p = p_bonf,
  holm_p = p_holm,
  bh_p = p_bh,

  raw_sig = p_values < 0.05,
  bonf_sig = p_bonf < 0.05,
  holm_sig = p_holm < 0.05,
  bh_sig = p_bh < 0.05
)

results
```

------------------------------------------------------------------------

### 2. ANOVA: comparing the means of 3+ groups

ANOVA: Analysis of Variance, variation between groups is compared to
variation within groups, this helps compare the means of the groups.

**Null hypothesis**: all group means are equal.

**Alternative hypothesis**: at least one group mean differs.

ANOVA tests all means at once, no need for testing correction here
(compared to multiple pairwise comparisons in t-tests)

Data set: We will use `data/catalyst_yield.csv` again, this time using
all four (A,B,C,D) catalysts. With ANOVA we can determine if different
catalysts produce different average yields, or are observed differences
just random variation within groups?

```{r}
yield_data <- read.csv("data/catalyst_yield.csv")
yield_data$catalyst <- factor(yield_data$catalyst)
head(yield_data)
```

**Task 1**: Visualise the four catalyst groups and their yield
percentage with a box plot.

```{r}
ggplot(yield_data, aes(x = catalyst, y = yield_pct, fill = catalyst)) +
  geom_boxplot() +
  geom_jitter() 
```

#### Assumptions

-   normality within each group (Shapiro-Wilk per group);
-   homogeneity of variances across groups (Levene's test).

**Task 2**: Test the assumptions above.

```{r}
# 2 ways 
#1st way filtering 
a_df <- yield_data %>%
  filter(catalyst == "A")

b_df <- yield_data %>%
  filter(catalyst == "B")

c_df <- yield_data %>%
  filter(catalyst == "C")

d_df <- yield_data %>%
  filter(catalyst == "D")

#shapiro.test(a_df$yield_pct)
#shapiro.test(b_df$yield_pct)
#shapiro.test(c_df$yield_pct)
#shapiro.test(d_df$yield_pct)

# 2nd way using pipes
yield_data %>% 
  group_by(catalyst) %>% 
  shapiro_test(yield_pct)
```

```{r}
levene_test(yield_data, yield_pct ~ catalyst)
```

**Task 3**: Run ANOVA using `aov`. Summarise the test results with
`summary()`.

```{r}
anova_fit <- aov(yield_pct ~ catalyst, yield_data)
summary(anova_fit)
```

**Let's look at the output**

1) What are residuals?

Answer: Residuals = within-group variation ("noise") → each `yield_pct`
is compared to its catalyst's mean yield.

2) What is the Sum Sq and Mean Sq here?

Answer: Sum of squares: measures total variability

catalyst: variability explained by differences between catalyst means,
residuals: variability within catalysts

Mean Sq = Sum Sq / DF (\~variance)

3) What is F-statistic and how does it relate to ANOVA?

Answer: F = Mean Sq (between groups) / Mean Sq (within groups)

If all catalysts had the same mean: between group variance ≈ within
group variance (F≈1)

if catalysts differ: between-group variance becomes much larger.

ANOVA indicates that at least one group differs from another, but not
which. To find out, we should run a post-hoc test.

**Task 5**: Run Tukey HSD using `TukeyHSD` on the fitted ANOVA model.
You can visualise the results of the Tukey HSD by using `plot()` .

```{r}
tukey_res <- TukeyHSD(anova_fit)
tukey_res
plot(tukey_res)
```

What can you conclude about the catalysts?

#### Kruskal-Wallis — non-parametric alternative for ANOVA

When ANOVA assumptions (normality, equal variance) are not met, we can
use `kruskal.test`. Instead of comparing means, it ranks all data points
and compares medians.

Kruskal-Wallis can be followed by the Dunn's Test for multiple
comparisons of the ranked data.

As the `airquality` data set is not normally distributed, and we would
like to compare more than 2 months, we can run Kruskal-Wallis:

```{r}
#kruskal.test(Ozone ~ Month, airquality)
dunn_test(airquality, Ozone ~ Month, p.adjust.method = "bonferroni")
```

Note: `dunn_test` already corrects p-values via `p.adjust.method`.
Whenever you do multiple pairwise comparisons, you must correct them.

#### Multi-factorial ANOVA

When more than one factor may affect the response, we should use a
multi-factorial ANOVA. The multi-factorial ANOVA enables to consider
individual effects of each factor (+ in formula), as well as the
interaction of factors (:) or both (\*). In the Appendix D you can see
it being applied on an example data set. In R, to assess the individual
effects as well as interaction effect of both factors, we would express
it this way: aov(outcome \~ factor1 \* factor2, data = dataframe).

------------------------------------------------------------------------

### 3. Correlation & regression

Theory: "slides/corr_simple_reg_26.pdf"

#### 3.1 Correlation

| Method | When | Notes |
|------------------------|------------------------|------------------------|
| **Pearson** | linear, both variables \~normal | sensitive to outliers |
| **Spearman** | monotonic, rank-based | robust to outliers, non-linear monotonic |
| **Kendall** | rank-based | better for small n / many ties |

Data set: we will use `data/fermentation.csv` which has data about
lactic acid concentration over time during fermentation for two starter
cultures. More exactly we will focus on the change of concentration over
time for Culture X.

```{r}
ferm <- read.csv("data/fermentation.csv")
head(ferm)

ferm_x <- ferm %>% filter(culture == "Culture_X")
```

**Task** **1**: Check the Pearson assumption (approximate bivariate
normality) by testing both variables (time and concentration).

```{r}
shapiro.test(ferm_x$time_h)
shapiro.test(ferm_x$lactic_acid_g_L)
```

**Task 2**: Run Pearson / Spearman / Kendall correlation. Pick the
appropriate `use = ...` argument to handle NAs.

```{r}
cor.test(ferm_x$time_h, ferm_x$lactic_acid_g_L, method="spearman")
cor(ferm_x$time_h, ferm_x$lactic_acid_g_L, method="spearman")
```

**Task 3:** Plot a scatter plot of concentration vs time for Culture X.

```{r}
ggplot(ferm_x, aes(x=time_h, y=lactic_acid_g_L)) + 
  geom_point() 
```

#### 3.2 Linear regression

**Task 1:** Fit a linear model (Culture X) with lactic acid
concentration being described by time (conc = intercept + slope \*
time). Inspect the summary using `summary()`.

```{r}
linear_model <- lm(lactic_acid_g_L ~ time_h, data = ferm_x)
summary(linear_model)
```

NB! The null hypothesis in linear regression tests for slope (i.e. "is
there a significant effect of time?") and intercept (i.e. "is the
predicted value at time 0 significantly different from 0?").

**Task 2:** Make sense of the summary output by answering:

1)  What are residuals in linear regression?

    Answer: Residuals are the differences between observed values and
    the values predicted by the regression model.

2)  What does the p-value for the slope mean?

    Answer: The p-value tests the hypothesis: H0: The true slope = 0 (no
    relationship between time and lactic acid)

    t-stat = slope / standard error of slope

3)  How are the degrees of freedom determined?

    Answer: Residual degrees of freedom is 28, since there are 30 data
    points - 2 parameters (intercept and slope).

**Task 3**: Let's plot the linear regression line on top (can be done
using `geom_abline()`) of the scatter plot created before.

```{r}
ggplot(ferm_x, aes(x=time_h, y=lactic_acid_g_L)) + 
  geom_point() +
  geom_abline(slope = 1.03525, intercept = 0.12249, color = "red")
  
```

#### 3.3 Standardisation with `scale()`

When predictors are on very different scales, coefficients are hard to
compare and may be numerically unstable. Standardising (like subtract by
mean, divide by SD) gives coefficients in units of *standard deviations*
of the predictor, making them comparable. See the Appendix C below for
an example. 1

### 4. Categorical data

Categorical data are often summarised as a **contingency (aka
association) table** (2 categorical variables, but can be beyond 2x2
table):

|              | Likes dogs | Does not like dogs |
|--------------|-----------:|-------------------:|
| Right-handed |         15 |                 25 |
| Left-handed  |         25 |                 45 |

#### 4.1 Chi-square test of independence

Tests whether there is a meaningful relationship between categorical
variables or whether observed data matches what you'd expect if there
were no effect/relationship.

Null hypothesis: there is no relationship between two categorical
variables.

$$
\chi^2 = \sum_{i=1}^{r} \sum_{j=1}^{c} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}
$$

where $r$, $c$ are the numbers of rows/columns, $O_{ij}$ are observed
and $E_{ij}$ are expected frequencies. Degrees of freedom: $(r-1)(c-1)$.

**Expected frequencies**:

$$
E_{ij} = \frac{(\text{Row Total})_i \times (\text{Column Total})_j}{\text{Grand Total}}
$$

**Assumptions**

-   large enough sample (all expected cell counts ≥ 5);
-   observations independent (each unit in exactly one cell).

```{r}
tbl <- matrix(c(15, 25, 25, 45), nrow = 2, byrow = TRUE,
              dimnames = list(handedness = c("Right", "Left"),
                              preference = c("Dogs",  "No dogs")))
tbl

row_tot <- rowSums(tbl)
col_tot <- colSums(tbl)
grand_tot <- sum(tbl)

expected <- outer(row_tot, col_tot) / grand_tot # outer() creates all row x column combination 
expected

chi <- chisq.test(tbl)
chi
```

**Task**: Using the `airquality` dataset, determine whether there is an
association between month and the temperature level. Categorise `Temp`
into ≥ 3 bins (e.g. low/medium/high with `cut()`) to satisfy the
Chi-square assumptions.

```{r}
temp_ranges <- cut(airquality$Temp, 3, labels = c("low", "medium", "high"))

months_factored <- factor(airquality$Month)

temp_tbl <- table(temp_ranges, months_factored)
temp_tbl

row_tot <- rowSums(temp_tbl)
col_tot <- colSums(temp_tbl)
grand_tot <- sum(temp_tbl)

expected <- outer(row_tot, col_tot) / grand_tot # outer() creates all row x column combination 
expected

chi <- chisq.test(temp_tbl)
chi
```

#### 4.2 Fisher's exact test

Use when:

-   any expected cell count is small (≤ 5);
-   total sample size is small (\< \~20) (but can also be used for
    larger sample sizes);

Fisher's test computes the exact p-value by creating all possible
combinations of 2×2 tables "at least as extreme" as the observed one,
given the fixed row and column totals (hence the term "exact").

Null hypothesis: categorical variables are independent/no association
(odds ratio = 1)

Alternative hypothesis: there is an association between the variables
(odds ratio ≠ 1) .

##### Classic example: tea tasting

Ronald Fisher's original example. A woman claims to distinguish whether
milk or tea was poured first; Fisher gave her 8 cups (4 of each). She
Fisher test with a one-sided test.

```{r}
tea_tasting <- matrix(c(3, 1, 1, 3),
                     nrow = 2,
                     dimnames = list(Guess = c("Milk", "Tea"),
                                     Truth = c("Milk", "Tea")))

tea_tasting
```

3 times milk was correctly identified as milk 3 times tea was correctly
identified as tea 1 misclassification each way.

We now test whether guessing is independent of truth (i.e., whether
people can actually distinguish milk-first vs tea-first pouring order).

```{r}
fisher.test(tea_tasting)
```

Odds ratio (OR) here refers: \
The observed imbalance in the sample suggests one group has about 6×
higher odds of a particular outcome.

→ People are more likely to correctly match categories than you’d expect
under independence.

P-values shows probability: "if there were really no association/no
ability to distinguish tea first or milk first, how often would we see a
table this extreme or more extreme?"

P = 0.49 → not that extreme

Even though OR looks large, small sample size can influence the p-value.

CI for odds ratios: wide range - from negative association to extremely
strong positive association (because the sample size is tiny)

**Task**: Let $X$ be a journal (*Mathematics Magazine* or *Science*),
and $Y$ be the topic of an article (math or biology). Suppose
*Mathematics Magazine* has 5 math articles and 1 biology article, and
*Science* has 1 math and 4 biology articles. Is there an association
between the journal and the topic? What does the output indicate?

```{r}
journal_tbl <- matrix(c(5, 1, 1, 4), byrow = TRUE,
                     nrow = 2,
                     dimnames = list(journal = c("MathMag", "Science"),
                                     topic = c("math", "biology")))

journal_tbl

fisher.test(journal_tbl)
```

------------------------------------------------------------------------

### 5. Wrap-up of Day 2

You can now:

-   correct p-values when running multiple tests;
-   compare 3+ groups parametrically (ANOVA + Tukey) and
    non-parametrically (Kruskal-Wallis + Dunn);
-   handle two factors at once with multi-factorial ANOVA;
-   quantify and model relationships between numeric variables
    (correlation, linear regression, standardisation);
-   analyse categorical data (Chi-square, Fisher's exact, odds ratio).

Please fill in this feedback form:<https://forms.gle/wXv72uFrck8a5xfF7>

**This marks the end of the course. Congrats!**

------------------------------------------------------------------------

## Appendix A: Polynomial regression (self-study)

When the relationship is non-linear in the predictor, fit a polynomial:

```{r}
data(iris)
linear_model <- lm(Sepal.Length ~ Sepal.Width, data = iris)
poly_model   <- lm(Sepal.Length ~ poly(Sepal.Width, 3, raw = TRUE), data = iris)

iris$pred_linear <- predict(linear_model, iris)
iris$pred_poly   <- predict(poly_model,   iris)

summary(linear_model)
summary(poly_model)

ggplot(iris, aes(Sepal.Width, Sepal.Length)) +
  geom_point(aes(colour = Species)) +
  geom_line(aes(y = pred_linear), colour = "red",   linetype = "dashed") +
  geom_line(aes(y = pred_poly),   colour = "green") +
  labs(title = "Linear vs polynomial fit") +
  theme_minimal()
```

------------------------------------------------------------------------

## Appendix B: Linear mixed models (self-study)

Mixed models handle **fixed effects** (what you care about, e.g.
treatment) together with **random effects** (variation you want to
account for but not estimate individually, e.g. experimental batch,
individual subjects).

Interactive intro: <http://mfviz.com/hierarchical-models/>

**When to use them**: when your data have hierarchical /
repeated-measures structure (multiple measurements per subject,
replicates within batches), and treating those measurements as
independent would inflate your false positive rate. Implementations:
`lme4::lmer()` and `lmerTest::lmer()`.

```{r}
# Example sketch (not run):
# library(lme4); library(lmerTest)
# lmer(lactic_acid_g_L ~ time_h + culture + (1 | replicate), data = ferm)
# Here `(1 | replicate)` says: each replicate gets its own random intercept.
```

This requires solid understanding of your study design — consult a
statistician for non-trivial cases.

## Appendix C: Example for standardisation

Let's take a look at a synthetic example where we have Target_Y that is
affected by both the Annual income (USD) and toxin levels in water
(mg/L) equally:

```{r}
# Set seed for reproducibility
set.seed(42)
n_samples <- 200

# Predictor A: Huge scale (e.g., Annual Income in USD)
# Mean = 70,000, SD = 15,000
Income_USD <- rnorm(n_samples, mean = 70000, sd = 15000)

# Predictor B: Tiny scale (e.g., Toxin levels in mg/L)
# Mean = 0.05, SD = 0.01
Toxin_mgL <- rnorm(n_samples, mean = 0.05, sd = 0.01)

# True relationship: Both variables contribute equally to Y's variance
# (15000 * 0.002 = 30) and (0.01 * 3000 = 30) -> Balanced impact
noise <- rnorm(n_samples, mean = 0, sd = 5)
Target_Y <- (0.002 * Income_USD) + (3000 * Toxin_mgL) + noise

# Combine into a data frame
df <- data.frame(Income_USD, Toxin_mgL, Target_Y)

# View the first few rows
head(df)
```

If we run unstandardised linear regression, we get:

```{r}
model_raw <- lm(Target_Y ~ Income_USD + Toxin_mgL, data = df)
print(model_raw)
```

Because the coefficients are completely dependent on the underlying
units (dollars vs. milligrams), it looks like the toxin has a massive
effect and income has zero effect. In reality, both have equal effect.

```{r}
model_standardized <- lm(Target_Y ~ scale(Income_USD) + scale(Toxin_mgL), data = df)
print(model_standardized)
```

With standardizing we are shifting from "What happens if we increase
income by \$1?", it asks: "What happens if we increase income by one
typical variation (Standard Deviation)?". This helps explain the
equality of the effect better.

------------------------------------------------------------------------

## Appendix D: Two-way ANOVA example

The built-in `ToothGrowth` dataset (The response is the length of
todontoblasts (cells responsible for tooth growth) in 60 guinea pigs by
vitamin C supplement type and dose):

```{r}
data(ToothGrowth)
?ToothGrowth
ToothGrowth$dose <- factor(ToothGrowth$dose)
head(ToothGrowth)
```

For running a multi-factorial ANOVA to test how the supplement type and
dose affects the outcome we can use

```{r}
# Two-way ANOVA with interaction
two_way <- aov(len ~ supp * dose, data = ToothGrowth)
summary(two_way)
```

The `*` includes both main effects of each factor and their interaction.

Here we are looking at if the supplement differ overall (main effect of
`supp`), does dose matter (main effect of `dose`), and does the effect
of dose and supplement together affect the outcome (interaction).

The main effect of `supp` asks: "If I ignore dose differences, do
supplements differ in mean outcome?"

The main effect of `dose` asks: "If I ignore supplement differences, do
doses differ in mean outcome?"

Use `+` if you only want main effects (independent contributions)

Use `:` for interaction (combined effect) only.
